Evaluation of Cache-Oblivious Data Structures [draft-1]

### 2008-05-26 paper/paper.tex
Revised paper.

\begin{abstract}
In modern computer hardware architecture, memory is organized in a hierarchy
consisting of several types of memory with different memory sizes, block
transfer sizes and access times. To reflect this situation more accurately
than tradition models, the cache-oblivious model has been proposed, and several
algorithms and data structures have been proposed that perform optimally in
this model.
This paper presents the results of an empirical evaluation of various data
structures in a realistic scenario, and gives insight into the applicability
of cache-oblivious data structure in practice.
\end{abstract}

\keywords{cache efficiency, locality of reference, algorithms, empirical evaluation}

@@

As a result, a modern computer system lacks a central memory storage with uniform performance characteristics. Instead, it employs a hierarchy of memory storage types. Figure \ref{fig-memhier} gives a typical example of such a hierarchy. The processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy. Memory is typically transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (,described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (which will also be described below) which does not allow parametrization with properties of the cache layout.
\end{list}

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Altough some research has been done for cache-oblivious data structures, they are seldom used in practice. The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.

\subsection{External Memory Model}
One of the earliest models to take the differences in memory access cost into account is the external memory model, described by Aggarwal and Vitter \cite{aggarwal1988ioc}. They make a distinction between internal memory (which is limited in size) and external memory (which is virtually unlimited). The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories; additionally, consecutive blocks can be transfered at reduced cost (so-called bulk transfer). Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy. In that case, the possibility of bulk transfer may have to be dropped.

@@

\subsection{Hierarchical Memory Model}
The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy contains more than just two levels. Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one. Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory address is described by a non-decreasing function of this address, which means that accessing data at higher addresses can be slower than at lower addresses. This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where, which is usually not the case in practice, making their model of little use in the design and evaluation of practical algorithms.

@@

Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration. The key advantage of this class of algorithms is that they are implicitly tuned to all levels in the memory hierarchy at once. It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.

@@

\section{Contributions}
Several cache-oblivious data structures and algorithms have been proposed. Complexity analysis shows that the proposed solutions are asymptotically optimal. However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often prefered that are easier to implement and require less memory).

@@

Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop. In their experimental results the cache-oblivious tree performs worse than the alternatives. Nevertheless, they conclude that \textit{"cache-obli\-vious data structures may have significant practical importance"}.

Askitis and Zobel \cite{askitis2005ccc} propose a way to make chained hash tables more cache-efficient, by storing bucket contents in contiguous memory instead of using a linked list. This means buckets have to be resized on insertions and deletions, but their experiments show a performance gain over traditional methods (especially when the hash table is heavily loaded).

@@

\subsection{State Space Search}
As the test application, that of a state space search has been chosen. This is a
suitable scenario for two reasons. First, it is commonly used as a practical
component of formal methods for software verification, and therefore good
algorithms are of great practical significance. Second, as we will explain
below, the performance of state space search algorithms depends for a large
part on the performance of the data structures that are used to implement
them; therefore, research into efficient data structures is of particular
interest to this application.

State space search is used to verify the correctness of software programs. For
this purpose, programs are first modeled using a formal language, that is also
used to specify properties of the program that should hold during its
execution. An executing program can be in a (possibly infinite) amount of states,
one of which is usually designated the \emph{initial state}. If a transition
from one state to another is possible (according to the rules of the formal
language used) the latter state is said to be a \emph{successor state} of the
former.
Generating the successors of a particular state is also called
\emph{expanding} the state. Of course, the model must include some form of
non-determinism to allow more than one successor to exist for a single state.
In practice, this non-determinism usually comes from processes that execute in
parallel, where the precise interleaving of the (observed) execution of
instructions in these processes is non-deterministic, unless synchronization
primitives (such as channels, semaphores, atomic execution blocks, et cetera)
are used to enforce a particular ordering.

@@

Depending on how states are inserted and extracted from the queue, the search
may be breadth-first (meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps) or depth-first (meaning that states that are generated most recently
by the successor function are expanded first). In the first case, the queue
operates by a first-in first-out principle, while in the second case the queue
works in a last-in first-out (stack) principle. In general, breadth-first
search requires a much larger queue capacity than depth-first search.

@@

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{spin}, a widely
used model checking tool. Spin uses a custom modeling language, called
PROMELA, to specify models and comes with a collection of example models that
are suitable for our experiment.

For the execution of these models we will use NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models.\cite{nipsvm} The NIPS VM is prefered to over the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

The NIPS VM represents program state as a binary string; the size of the state
depends on the number of active processes in the program, and is typically in
the order of a few hundred bytes.

\subsection{Data Structures}
In the introduction we identified three classes of data structures.
For our evaluation we have implemented a single representative data structure
for each class:
\begin{itemize}
\item \textbf{Cache-unaware}: hash tables. Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
\item \textbf{Cache-aware}: B-trees. The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
\item \textbf{Cache-oblivious}: the B-tree-like data structure proposed by Bender, Duan, Iacono and Wu. Since it provides functionality comparable to that of a normal B-tree, this seems like a fair candidate for a comparison. For brevity, this data structure will be refered to as a \emph{Bender set}.
\end{itemize}

@@

In contrast, both hash tables and B-trees are widely used and there are several
high quality implementations available as software libraries. It is, however,
undesirable to use existing libraries for a fair comparison, for two reasons.

First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations. Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not. This may give the existing libraries an unfair advantage.

@@

\subsubsection{Set Operations}
Dynamic set data structures may support various operations, such as adding and removing elements, testing if a value is an element of the set, finding elements nearby, succeeding or preceding a given value, counting elements in a range, et cetera. However, our state space search algorithm only needs two operations: insertion of new elements (in a set that is initially empty) and testing for the existence of elements. In fact, these two operations can be combined into a single operation. We call this operation \verb#insert(S,x)#. If an element $x$ does not exist in $S$, then $insert(S,x)$ inserts it and returns $0$. If $x$ is already an element of $S$, $insert(S,x)$ returns $1$. The inner loop of our state space algorithm can now be written as follows:
\begin{verbatim}
    for (State s : successors(state)) {
        if (generated.insert(s) == 0) {
            queue.insert(s);
        }
    }
\end{verbatim}

@@

The use of memory mapped files means that the operating system is in charge of
when and which pages are transfered between main memory and disk storage; the
application assumes no control over this policy. This is in accordance with the
cache-oblivious model, that assumes no control over the cache is possible.

\subsubsection{Hash Table Implementation}
Hash tables are widely used as a fast (O($1$)) data structure especially when
used for in-memory storage. In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map key values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existance of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

@@

B-trees are similar to other search tree data structures. A defining property
is that they organize data in pages of a fixed size, each containing several
values, which makes them especially useful for purposes of on-disk storage
where reading and writing data in a few large blocks is relatively cheap
compared to accessing several smaller blocks.

B-trees work by keeping as much values as possible, in order, in a single
page. Pages are ordered in a tree structure, where interior nodes store
pointers to child nodes. Pointers originate at a pair of adjacent values,
and point to a subtree containing only values that fall in between those
values.

When a new value does not fit in the page where it is to be inserted,
the page is split into two pages; the median value from the old page is
moved to the parent page, and two new pages (containing the values less
than respectively greater than the median) are created. When the top-most
page needs to be split, a new (empty) top-level page is created and the height
of the B-tree is increased by one.

B-trees are well suited to storing variable length values, as pages can
store many values as long as the combined size of the values and pointers
does not exceed the pagesize.

@@

Note that there is no obvious way to store values that are larger than the
size of a single page. Production-quality implementations handle this case
separately, but our implementation does not support this scenario and
requires stored values to be smaller than the pagesize. Additionally,
deletion of items has not been implemented.

\subsubsection{Bender Set Implementation}
The first implementation challenge for the Bender set is that the description
given by Bender et al. assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of spaces wasted
remains below 50\% while still allowing values of various sizes to be inserted.

The implementation of the Bender set consists of two parts: a sparse array
storing the values in-order, and a binary tree stored in van Emde Boas
layout (similar to the one proposed by Prokop). Both the tree and the array
may contain special \emph{empty values}. A depiction of a Bender set storing
four values (with a capacity for eight) is given in Figure \ref{fig-bender-set}.

@@

A Bender set is created with a capacity $C$ that is a power of two.
Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size $log_2 C$
(rounded up to the next power of 2).
For each level, a maximum density is given, with the lowest level having density 1,
the highest level having density $\delta$ (these densities are interpolated linearly
to obtain the density for the intermediary levels).

The top-level density must be a number between $\frac{1}{2}$ and $1$; the optimal value
depends on practical circumstances such as available memory and the relative cost of
rebuilding the data structures. Whenever the number of elements stored in
a window (the window's \emph{population count}) exceeds the size of the window
multiplied by its maximum density, the window is said to be \emph{overflowing}.
Overflows are resolved when a higher-level window is \emph{rebalanced},
meaning that the values in the window are redistributed evenly over the window,
which will cause some of the values to be moved out from the lower-level window
which is too full.

Now, to insert a value, the binary tree is used to find a position in the array
where the value may be inserted (just before its successor). Then, the windows
overlapping this position are considered from bottom to top, and the first
window that can support another element without overflowing is selected, and
rebalanced (thereby resolving the overflow in all lower level windows).

In the case that all windows overlapping the goal position are overflowing, the
capacity of the data structure is doubled, a new top level is created and then
the entire array is rebalanced. This is how the set grows when the number of
elements stored comes too close to the storage capacity.

According to this description (which follows the paper by Bender et al) a window
is rebalanced every time an element is inserted. As an optimization, our
implementation does not always rebalance a window. When there is free space
between the successor and predecessor of the value to be inserted, the value
is simply inserted in the middle. In our experiments, this optimization
increased performance somewhat.

One thing that is not specified in the paper by Bender et al, is how the
population count for windows is kept. One simply option is to keep no such
information and simply scan part of the array whenever a population count is
required. Another extreme is to keep population counts for all windows on all
levels, and update these counts whenever values are inserted or moved. In
our experiments a compromise seemed to work best: keep population counts only
for the lowest-level windows, and recompute the counts for higher-level windows
when required. This saves a lot of recomputation, while keeping the additional
costs of updating low.

@@

\subsection{Measurements}
Measuring the memory usage of a process can be done by retrieving how much of the
address space has been allocated by a process; typically called the
\emph{virtual set size}.
Since a single page can be mapped into memory more than once (either in multiple
processes, or in the same process) this is not exactly equal to the amount of
memory allocated for the process, but since the majority of the memory used by
the data structures is mapped at exactly one location, it is good enough as
an indication of memory use.

There are several ways of measuring the time a process takes to execute.
The simplest is counting the number of seconds elapsed since the start of the
program (called "wall clock time"). This has the disadvantage that concurrently
executing processes affect the runtime of the process being measured, so on a
busy system the reported runtime may vary with different executions.

A different way to measure time is using the statistics the kernel collects
of how many seconds the CPU spends executing instructions for the process
(both in user space and in kernel space). These values are independent of
what other processes are doing, which makes them more consistent across
multiple executions. However, these values do not account for the time the
system spends waiting (for example, for a page to be swapped in from disk)
or the time spend by other processes on behalf of the executing process
(for example, the kernel swap daemon on Linux). Since these factors may have
a large effect on the total running time of the algorithm, it is undesirable
to leave these out.

Therefore, we decided to measure wall clock time. In order to deal with
variations across multiple executions, each experiment is run three times,
and the median value is used for further analysis.

\subsubsection{Framework Overhead}
Finally, there is another factor that must be taken into account: the test
framework uses several components (such as the NIPS VM and a queue) that are
not being evaluated, yet which do contribute to the memory use and runtime
of the process. In order to discount these factors, tests are performed using
a mock set implementation that gives a near-minimal overhead. This works
by first running with a real set implementation and logging all results
(the return values of the \verb#insert(S,x)# calls) to a file, and then
running a second time while reading the stored values. As a result, the mock
set implementation has very little memory and runtime overhead.

@@

The following models are used for benchmarking:
\begin{itemize}
\item\verb#eratosthenes# is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

\item\verb#leader2# is a leader election algorithm with a configurable number
of processes competing for leadership ($N$). Since a constant number of
processes will be created and these processes cannot make progress until all
of them are created, almost all states have the same size.

\item\verb#peterson_N# is a Peterson's solution to the multi-process mutual
exclusion problem (using global variables instead of channels for
communication) with a configurable number of processes ($N$). The processes
are created atomically, so the state size is constant after initialization.
\end{itemize}

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Case} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the properties of the execution of the model. Note that the number of transitions
relative to the number of iterations indicates the percentage of insertion operations
that results in a new element being added to the set.

>>>>

\begin{abstract}
In modern computer hardware architecture memory is organized in a hierarchy
consisting of several types of memory with different memory sizes, block
transfer sizes and access times. To reflect this situation more accurately
than traditional models, the cache-oblivious model has been proposed, and
several algorithms and data structures have been developed that perform
optimally in this model.
This paper presents the results of an empirical evaluation of several data
structures in a realistic scenario, and gives insight into the applicability
of cache-oblivious data structure in practice.
\end{abstract}

\keywords{cache efficiency, locality of reference, algorithms}

@@

As a result, a modern computer system lacks a central memory storage with uniform performance characteristics. Instead, it employs a hierarchy of memory storage types. Figure \ref{fig-memhier} gives a typical example of such a hierarchy. The processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transferred upward through the memory hierarchy. Memory is typically transferred in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existance of a memory hierarchy, and may therefore not perform optimally in practice.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
\end{list}

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Altough some research has been done on cache-oblivious data structures, they are seldom used in practice. The main contribution of this paper is to give some insight in the practical merit of cache-oblivious data structures.

\subsection{External Memory Model}
One of the earliest models to take the differences in memory access cost into account is the external memory model, described by Aggarwal and Vitter \cite{aggarwal1988ioc}. They make a distinction between internal memory (which is limited in size) and external memory (which is virtually unlimited). The external memory is subdivided into blocks of a fixed size and only entire blocks of data can be transferred between the two memories; additionally, consecutive blocks can be transferred at reduced cost (so-called bulk transfer). Although Aggarwal and Vitter focus on magnetic disk as a storage medium for external memory, the model can be generalized to apply to every pair of adjacent levels in the memory hierarchy. In that case, the possibility of bulk transfer may have to be dropped.

@@

\subsection{Hierarchical Memory Model}
The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy contains more than just two levels. Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one. Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory addresses is described by a non-decreasing function of these addresses, which means that accessing data at higher addresses can be slower than at lower addresses. This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where and that data at all locations can be accessed directly, which is usually not the case in practice. As a result, the applicability of their model to the design and evaluation of \emph{practical} algorithms is limited.

@@

Algorithms that perform optimally in this model are called ``cache-oblivious'' and they distinguish themselves from cache-aware algorithms in that they cannot rely on knowledge of the block size or other specific properties of the cache configuration. The key advantage of this class of algorithms is that, even though they are defined in a two-level memory model, they are implicitly tuned to all levels in the memory hierarchy at once. It has been conjectured that these algorithms may therefore perform better than algorithms that are tuned to a specific level in the hierarchy only.

@@

\section{Contributions}
Several cache-oblivious data structures and algorithms have been proposed. Complexity analysis shows that the proposed solutions are asymptotically optimal. However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often preferred that are easier to implement and require less memory).

@@

Several data structures were proposed by Rahman, Cole and Raman \cite{rahman2001opd} amongst them a cache-oblivious exponential search tree with a similar structure as the static search tree proposed by Prokop. In their experimental results the cache-oblivious tree performs worse than the (non-oblivious) alternatives. Nevertheless, they conclude that \textit{``cache-obli\-vious data structures may have significant practical importance''}.

Askitis and Zobel \cite{askitis2005ccc} propose a way to optimize seperate-chaining hash tables for cache-efficiency by storing the linked lists that contain the contents for a bucket in contiguous memory. Their experiments show a performance gain over traditional methods, especially when the hash table is heavily loaded.

@@

\subsection{State Space Search}
The test application that we use to gather performance data is an implementation
of a state space search algorithm.
This is a suitable scenario for two reasons.
First, it is commonly used as a practical
component of formal methods for software verification, and therefore good
algorithms are of great practical significance. Second, as we will explain
below, the performance of state space search algorithms depends for a large
part on the performance of the data structures that are used to implement
them; therefore, research into efficient data structures is of particular
interest to this application.

State space search can be used to verify the correctness of software programs. For this purpose, programs are first modeled using a formal language, that is also used to specify properties of the program that should hold during its
execution. An executing program can be in a (possibly infinite) amount of states,
one of which is usually designated the \emph{initial state}. If a transition
from one state to another is possible (according to the rules of the formal
language used) the latter state is said to be a \emph{successor state} of the
former.
Generating the successors of a particular state is also called
\emph{expanding} the state. Of course, the the execution model
must include some form of non-determinism to allow more than one successor to
exist for a single state.
In practice, this non-determinism usually comes from processes that execute in
parallel, where the precise interleaving of the execution of
instructions in these processes is non-deterministic, unless synchronization
primitives (such as channels, semaphores, atomic execution blocks, et cetera)
are used to enforce a particular ordering.

@@

Depending on how states are inserted and extracted from the queue, the search
may be breadth-first (meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps) or depth-first (meaning that states that are generated most recently
by the successor function are expanded first). We will use breadth-first
search, which means the queue operates by a first-in first-out principle.

@@

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{spin}, a widely
used model checking tool. Spin uses a custom modeling language (called
PROMELA) to specify models and is distributed with a collection of example
models that are suitable for our experiments.

For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models.\cite{nipsvm} The NIPS VM is preferred to the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

The NIPS VM represents program state as a binary string; the size of the state
depends amongst other on the number of active processes in the program (which may change over the execution of the program). A typical state size is in the order of a few hundred bytes.

\subsection{Data Structures}
In the introduction we identified three classes of data structures.
For our evaluation we have implemented a single representative data structure
for each class:
\begin{itemize}
\item \textbf{Cache-unaware}: hash tables. Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
\item \textbf{Cache-aware}: B-trees. The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
\item \textbf{Cache-oblivious}: the data structure proposed by Bender, Duan, Iacono and Wu. Since it provides functionality comparable to that of a B-tree, this seems like a fair candidate for a comparison. For brevity, this data structure will be referred to as a \emph{Bender set}.
\end{itemize}

@@

In contrast, both hash tables and B-trees are widely used and there are several
high quality implementations available as software libraries. It is, however,
undesirable to use existing libraries for a fair comparison, for two reasons. First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations. Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not. This may give the existing libraries an unfair advantage.

@@

\subsubsection{Set Operations}
Dynamic set data structures may support various operations, such as adding and removing elements, testing if a value is an element of the set, finding elements nearby, succeeding or preceding a given value, counting elements in a range, et cetera. However, our state space search algorithm only needs two operations: insertion of new elements (in a set that is initially empty) and testing for the existence of elements. In fact, these two operations can be combined into a single operation. We call this operation \verb#insert(S,x)#. If an element $x$ does not exist in $S$, then \verb#insert(S,x)# inserts it and returns $0$. If $x$ is already an element of $S$, \verb#insert(S,x)# returns $1$ and no modifications are made. The inner loop of our state space algorithm can then be rewritten as follows:
\begin{verbatim}
    for (State s : successors(state)) {
        if (generated.insert(s) == 0) {
            queue.insert(s);
        }
    }
\end{verbatim}

@@

The use of memory mapped files means that the operating system is in charge of
when and which pages are transferred between main memory and disk storage; the
application assumes no control over this policy. This is in accordance with the
cache-oblivious model, that assumes no control over the cache is possible.
Additionally, memory mapped files bypass the C library's allocation functions
and give the implementor complete control over how data is ordered in (virtual)
memory.

\subsubsection{Hash Table Implementation}
In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map key values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existance of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

@@

B-trees are similar to other search tree data structures in the sense that
they store ordered values hierarchically in such a way that each subtree store
values that are close together.
A key property of B-trees is that they do not fix the number of children per node, but instead organize data in pages of a fixed size, each containing
as many values as will fit. This makes them especially suitable for purposes
of on-disk storage where reading and writing data in a few large blocks is
relatively cheap compared to accessing several smaller blocks.

B-tree pages are ordered in a tree structure. Values in a page are stored
in order, and pointers are added that originate between a pair of adjacent
values, and point to a subtree containing only values that fall in between those
values. The resulting tree is typically very shallow, which is benefical,
as the number of pages that need to be retrieved (worst case) is equal to
the height of the tree.

When a new value does not fit in the page where it is to be inserted,
the page is split into two pages; the median value from the old page is
moved to the parent page, and two new pages (containing the values less
than respectively greater than the median) are created. When the top-most
page needs to be split, a new (empty) top-level page is created and the height
of the B-tree is increased by one. As a result, all leaf nodes in a B-tree are
at the same depth, and all pages stored are at least half full.

@@

B-trees can easily support the insertion of variable-length values, as long
as individual values fit in the size of a single page. However, there is no
obvious way to store values that are larger than the size of a single page.
Production-quality implementations handle this case separately. Our our
implementation does not support this requires stored values to
be smaller than the pagesize. Additionally, deletion of items has not been
implemented.

\subsubsection{Bender Set Implementation}
The first implementation challenge for the Bender set is that the description
given by Bender et al. assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of spaces wasted
remains below 50\% while still allowing values of various sizes to be inserted.
In the following description, we will assume that values are of a fixed size.

The implementation of the Bender set consists of two parts: a sparse array
storing the values in order, and a binary tree stored in van Emde Boas
layout (similar to the one proposed by Prokop) that is used to find values
efficiently.
Both the tree and the array may contain special \emph{empty values}.
A depiction of a Bender set storing four values (with a capacity for eight) is
given in Figure \ref{fig-bender-set}.

@@

A Bender set is created with a capacity $C$ that is a power of two.
Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size
$log_2 C$ (rounded up to the next power of 2).
For each level a maximum \emoph{density} is chosen, with the lowest level having
density 1, the highest level having density $\delta$, and the density for
the intermediary levels obtained by interpolating linearly between 1 and
$\delta$.

The top-level density must be a number between $0$ and $1$; the optimal value
depends on practical circumstances such as available memory and the relative
cost of rebuilding the data structures.
Whenever the fraction of non-empty values stored in a window (the window's
\emph{population count} divided by the size of the window) exceeds its maximum
density, the window is said to be \emph{overflowing}.
Overflows are resolved when a higher-level window is \emph{rebalanced},
meaning that the values in the window are redistributed evenly over the window,
which will cause some of the values to be moved out of the lower-level windows
which are too full.

To insert a value, the binary tree is used to find a position in the array
where the value may be inserted (just before its successor). Then, the windows
overlapping this position are considered from bottom to top, and the first
window that can support another element without overflowing is selected, and
then rebalanced (thereby resolving the overflow in all lower level windows).

In the case that all windows overlapping the target position would overflow upon
insertion of the new element, the capacity of the data structure is doubled,
a new top level is created and then the entire array is rebalanced. This is how
the set grows when the number of elements stored comes too close to the storage
capacity. Since changing the capacity causes all values in the array to be moved
and requires that the entire index tree is recreated, this is an expensive
operation; however, it occurs only infrequently.

According to this description (which follows the paper by Bender et al) a window
is rebalanced every time an element is inserted. As an optimization, our
implementation does not always rebalance a window. When there is free space
between the successor and predecessor of the value to be inserted, the value
is simply inserted in the middle. In our experiments, this optimization
yielded an reduction in execution time.

One thing that is not specified in the paper by Bender et al, is how the
population count for windows is kept. The simplest option is to keep no such
information and simply scan part of the array whenever a population count is
required. Another extreme is to keep population counts for all windows on all
levels, and update these counts whenever values are inserted or moved. In
our experiments a compromise seemed to work best: keep population counts only
for the lowest-level windows, and recompute the counts for higher-level windows
when required. This prevents a lot of recomputation, while keeping the additional
costs of updating low.

@@

\subsection{Measurements}
Measuring the memory usage of a process can be done by retrieving how much of the
address space has been allocated by a process; called the \emph{virtual set
size}. This does not necessarily correspond one-to-one with memory being allocated for the process, but since the majority of the memory used by the
data structures is mapped at exactly one location, this metric is suitable
for our experiments.

There are several ways of measuring the time a process takes to execute.
The simplest is counting the number of seconds elapsed since the start of the
program (called \emph{wall clock time}).
This has the disadvantage that concurrently executing processes affect the runtime
of the process being measured, so on a busy system the reported runtime may vary
over different executions.

A different way to measure time is using the statistics the kernel collects
of how many seconds the processor spends executing instructions for the process
(both in user space and in kernel space). These values are independent of
what other processes are doing, which makes them more consistent across
multiple executions. However, these values do not account for the time the
system spends waiting (for example, for data to read from disk)
or the time spent by other processes on behalf of the executing process
(for example, the kernel swap daemon on Linux). Since these factors may have
a large effect on the total running time of the algorithm, it is undesirable
to leave these out.

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run three times, and the median
value is used for further analysis.

\subsubsection{Framework Overhead}
Finally, there is another factor that must be taken into account: the test
framework uses several components (such as the NIPS VM and a queue) that are
not being evaluated, yet which do contribute to the memory use and runtime
of the process. In order to discount these factors, tests are performed using
a mock set implementation that gives a near-minimal overhead. This works
by first running with a real set implementation and logging all results
(the return values of the \verb#insert(S,x)# calls) to a file, and then
running a second time while reading the stored values. In the second run,
the mock set implementation does not have to actually store or retrieve
data and consequently has negligible memory and runtime overhead.

@@

The following models are used for benchmarking:
\begin{itemize}
\item\textbf{Eratosthenes} is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers upto a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

\item\textbf{Leader2} is a leader election algorithm with a configurable number
of processes competing for leadership ($N$). Since a constant number of
processes will be created and these processes cannot make progress until all
of them are created, almost all states have the same size.

\item\textbf{Peterson\_N} is Peterson's solution to the multi-process mutual
exclusion problem (using global variables instead of channels for
communication) with a configurable number of processes ($N$). The processes
are created atomically, so the state size is constant after initialization.
\end{itemize}

\begin{table}
\begin{center}
\begin{tabular}{ l l l l }
\hline
\textbf{Model} & \textbf{Parameters} & \textbf{Iterations} & \textbf{Transitions} \\
\hline
Eratosthenes & $N=40$ &  1,019,960 &  4,923,218 \\
Leader2      & $N=5$  &  5,950,945 & 23,856,363 \\
Peterson\_N  & $N=4$  & 10,000,000 & 37,434,411 \\
\hline
\end{tabular}
\caption{Properties of test cases used}
\label{tab-cases}
\end{center}
\end{table}

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the resulting properties of the state space search.
Note that the number of transitions relative to the number of iterations indicates
the percentage of insertion operations that results in a new element being added
to the set, which ranges approximately from 4 to 2.7 non-updating queries per
insertion.
